---
title: "README"
author: "Matt Kaye"
date: "6/7/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

rm(list = ls())

library(tidyverse)
library(rprojroot)
library(data.table)
library(lubridate)

PROJECT_ROOT <- find_root('CitiBike.Rproj')
```

## Exploring the New York City CitiBike data

This project explores the CitiBike program data from New York City.

### Data

The data comes from the [CitiBike program](https://www.citibikenyc.com/system-data) directly, and is publicly available for download.

Below is a glimpse at the structure of the data

```{r}
fread(paste(PROJECT_ROOT, '/data/citibike_dat_1.csv', sep = "")) %>%
  mutate(starttime = starttime %>% as_datetime(),
         stoptime = stoptime %>% as_datetime(),
         bikeid = bikeid %>% as.factor()) %>%
  sample_n(size = 100) %>%
  glimpse()
```


## Plot Examples

#### Visualizing Commute Patterns – Heatmap of start and end stations by time of day

<img src="https://raw.githubusercontent.com/mrkaye97/CitiBike/master/viz/commutes.svg">

#### Coronavirus Impact on Ridership

<img src="https://raw.githubusercontent.com/mrkaye97/CitiBike/master/viz/coronavirus-overall-impact.svg">
