---
title: "README"
author: "Matt Kaye"
date: "6/7/2020"
output: github_document
---

```{r, echo = F}
knitr::opts_chunk$set(echo = T)
```

```{r setup}
suppressPackageStartupMessages(library(tidyverse)) 
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(rprojroot))
suppressPackageStartupMessages(library(lubridate))

PROJECT_ROOT <- find_root('CitiBike.Rproj')
```


## Exploring the New York City CitiBike data

This project explores the CitiBike program data from New York City.

A few notes:

* The full project can be found [Here](https://mrkaye97.github.io/CitiBike/).
* All R code is in `~/CitiBike/R`
* All python code is in `~/CitiBike/python`
* Some data used in certain files can be found in `~/CitiBike/data`, but much is pulled directly from online sources.


### Setup

The project requires no explicit setup. All data is either included in this repo or comes directly from the CitiBike website.

A handful of R and Python packages are needed to run the project. See the bottom of this README for a list.

### Data

The data comes from the [CitiBike program](https://www.citibikenyc.com/system-data) directly, and is publicly available for download.

Below is a glimpse at the structure of the data

```{r}
fread(paste(PROJECT_ROOT, '/data/citibike_dat_1.csv', sep = "")) %>%
  mutate(starttime = starttime %>% as_datetime(),
         stoptime = stoptime %>% as_datetime(),
         bikeid = bikeid %>% as.factor()) %>%
  sample_n(size = 100) %>%
  glimpse()
```


## Plot Examples

#### Visualizing Commute Patterns – Heatmap of start and end stations by time of day
<img src="https://raw.githubusercontent.com/mrkaye97/CitiBike/master/viz/commutes.svg">

#### Visualizing Rides Broken Down by Year, Day of the Week, and Time of Day
<img src="https://raw.githubusercontent.com/mrkaye97/CitiBike/master/viz/coronavirus-and-day-dsn.svg" width="980" height="1000">

